---
title: "AMA Lab 5: Local Poisson Regression"
author: "Cecilia, Rudio, Gerard"
date: "2023-11-18"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(sm)
countries<-read.csv2(file="HDI.2017.subset.csv",row.names = 1)
attach(countries)
head(countries)
```

As a quick summary of the data, the mean life expectancy of the countries in this dataset is 71.85 being 73.40 as the median and for female 74.28 higher than the lif.expec of males, 69.45. The mean infant mortality rate is 23.26.

```{r}
summary(countries)
```
```{r}
plot(countries[,2:7])
```
```{r}
ind.agr<-(Agric.employ.. < median(Agric.employ..))
```


## 1. Bandwidth choice for the local Poisson regression

To adapt the function loglik.CV to we need to use sm.poisson to predict and modify the calculus of the log-likelihood according to this expressions :

\[ll_{CV}(h)
=\frac{1}{n}\sum_{i=1}^{n} \log\left( \widehat{\Pr}_h^{\tiny(-i)}(Y=y_i|X=x_i)\right),
\]

where $\widehat{\Pr}_h^{\tiny(-i)}(Y=y_i|X=x_i)$ is an estimation of
\[
\Pr(Y=y_i|X=x_i)=e^{-\lambda_i} \frac{\lambda_i^{y_i}}{y_i!},
\]

with $\lambda_i=E(Y |X = xi )$ estimated with sm.poisson.


```{r cars}

loglik.CV.pois <- function(x,y,h){
  n <- length(x)
  pred <- sapply(1:n, 
      function(i,x,y,h){
         sm.poisson(x=x[-i],y=y[-i],h=h,eval.points=x[i],display="none")$estimate
      },   x,y,h)
  return(-sum(log(exp(-pred)*pred^(y)/factorial(y)))/n)
}

h.cv.sm.pois <- function(x,y,rg.h=NULL,l.h=10,method=loglik.CV.pois){
   cv.h <- numeric(l.h)
   if (is.null(rg.h)){
      hh <- c(h.select(x,y,method="cv"),
              h.select(x,y,method="aicc"))#,hcv(x,y)
      rg.h <- range(hh)*c(1/1.1, 1.5)
   }
   i <- 0
   gr.h <- exp( seq(log(rg.h[1]), log(rg.h[2]), l=l.h))
   for (h in gr.h){
      i <- i+1
      cv.h[i] <- method(x,y,h)
   }
   return(list(h = gr.h, 
               cv.h = cv.h, 
               h.cv = gr.h[which.min(cv.h)]))
}
```

## 2. Local Poisson regression for Country Development Data

```{r}
le.fm.r <- round(le.fm)
```

Here it is stated the code is performing cross-validation for Poisson kernel density estimation on a new dataset. Performs cross validation on the dataset provided with the values le.fm.r.

```{r}
h.CV.pois.lefm <- h.cv.sm.pois(Life.expec,le.fm.r,rg.h=c(1,10), method = loglik.CV.pois)
```

The plotting includes a line plot of cross validated log-likelihood against bandwidth, being x axis the h values (bandwidth) and y axis the los-likelihood cross-validated values. In addition, a red point on the plot is added showingthe optimal h for the cross-validated log-likelihood.

```{r}
plot(h.CV.pois.lefm$h, h.CV.pois.lefm$cv.h, type = "l", 
     xlab = "Bandwidth (h)", ylab = "Cross-validated Log-Likelihood",
     main = "Cross-validated Log-Likelihood vs Bandwidth")
points(h.CV.pois.lefm$h, h.CV.pois.lefm$cv.h)
points(h.CV.pois.lefm$h.cv, min(h.CV.pois.lefm$cv.h), col = "red", pch = 16)
```

Finally, the poisson regression using the bandwidth obtained from corss-validation. A line is added showing the local Poisson regression curve to the plot using the fitted values.

```{r}
pois_fit <- sm.poisson(Life.expec,le.fm.r, h = h.CV.pois.lefm$h.cv,col='blue')
```

