---
title: "AMA7 : GAM fits for hirsutism data"
author: "Cecilia, Rudio and Gerard"
date: "2023-12-06"
output: html_document
---


```{r, echo=FALSE}
library(mgcv)
hirs <- read.table("hirsutism.dat",header=T, sep="\t",fill=TRUE)

hirs$Treatment <- as.factor(hirs$Treatment)
```


```{r, echo=FALSE}
summary(hirs)
```


```{r , echo=FALSE}
attach(hirs)

boxplot(hirs[,2:5])

par(mfrow=c(2,2))
boxplot(hirs[,2]~Treatment,ylim=c(0,30), main=names(hirs)[2], xlab="Treatment")
boxplot(hirs[,3]~Treatment,ylim=c(0,30), main=names(hirs)[3], xlab="Treatment")
boxplot(hirs[,4]~Treatment,ylim=c(0,30), main=names(hirs)[4], xlab="Treatment")
boxplot(hirs[,5]~Treatment,ylim=c(0,30), main=names(hirs)[5], xlab="Treatment")
par(mfrow=c(1,1))

par(mfrow=c(2,2))
boxplot(hirs[Treatment==0,2:5],ylim=c(0,30), main="Treatment 0")
boxplot(hirs[Treatment==1,2:5],ylim=c(0,30), main="Treatment 1")
boxplot(hirs[Treatment==2,2:5],ylim=c(0,30), main="Treatment 2")
boxplot(hirs[Treatment==3,2:5],ylim=c(0,30), main="Treatment 3")
par(mfrow=c(1,1))
```

The following boxplots indicate the distribution of the records for thr FG values at different times and for different treatments. For instance, as a quick summary, for early stages of FG values there is a larger quantity of records for treatment. The lowest median lies for treatment 3 and the largest whiskers for treatment 1 and 3. Both treatment 2 and 3 include outliers. For FG values for month 3 all box heights appear very equal between treatments being treatment 2 the one with the lowest median and treatment 1 the one with the longest whiskers. Box heights differ for FG values for month 6 being treatment 3 the one with the highest box height and 2 the lowest. Treatment 1 and 2 both appear to have outliers and treatment 0 appears to have the highest mean above all treatments. Finally, for FG values for month 12 the highest median belongs to treatment 0 but the outliers lie in treatmetn 1 and 2. The highest variability for values ofr FGm12 lie for treatment 3. For the second batch of boxplots, it can be seen how the PG values decrease as month 12 approaches for all values, being Treatment 2 the one with the highest decrease between month 0 and 12.
##Modeling FGm12

First a model with all the variables and without smoothing is fitted.

```{r}
mod1 <- gam(FGm12~SysPres+DiaPres+weight+height+FGm0+Treatment, data = hirs)
summary(mod1)
```

Looking at the summary, it can be seen that the only significant variable is FGm0. All the other variables, SysPres, DiaPres,weight, and height are not significant. Next the same model but with all the variables smoothed is fitted:

```{r}
mod1s <- gam(FGm12~s(SysPres)+s(DiaPres)+s(weight)+s(height)+s(FGm0)+Treatment, data = hirs)
summary(mod1s)
plot(mod1s,pages=1,residuals=TRUE, shade=TRUE, cex=2, lwd=2)
```

As a result of this model it can be seen that again there isthe FGm0 variable significant, and also the intercept.

```{r}
par(mfrow = c(2,2))
gam.check(mod1s)
```

It can be clearly seen that the residuals have a parabolic shape, therefore sqrt the FGm12 in order to see if the residuals distribution improves is a choice.


```{r}
mod1sqrts <- gam(sqrt(FGm12)~s(SysPres)+s(DiaPres)+s(weight)+s(height)+s(FGm0)+Treatment, data = hirs)
summary(mod1sqrts)
par(mfrow = c(2,2))
gam.check(mod1sqrts)
```

Looking at the plots for the sqrt model,it can be seen an improvement in the plots. First, the residuals follow a normal distribution. The parabolic pattern there was in the resids vs linear pred is lost.

```{r}
mod1sqrts.1 <- gam(sqrt(FGm12)~s(FGm0)+Treatment, data = hirs)
summary(mod1sqrts.1)
plot(mod1sqrts.1,pages=1,residuals=TRUE, shade=TRUE, cex=2, lwd=2)
```

A new model is fitted including DiaPres variables as it was one of the highest edf values in the smoothed initial model, for sake of argumentation.

```{r}
mod1sqrts.3 <- gam(sqrt(FGm12)~s(FGm0)+s(DiaPres)+Treatment, data = hirs)
summary(mod1sqrts.3)
plot(mod1sqrts.3,pages=1,residuals=TRUE, shade=TRUE, cex=2, lwd=2)
```

As a next step the interaction between the variables and the treatment is added

```{r}
mod1sqrts.2 <- gam(sqrt(FGm12)~s(FGm0,by=Treatment), data = hirs)
summary(mod1sqrts.2)
par(mfrow = c(2,2))
gam.check(mod1sqrts.2)
```

It can be seen how we are loosing the normality in the residuals. The smooth Systolic Preassure is not significant, so lets try to remove it and look at how it affects the model.

```{r}
mod1sqrts.4 <- gam(sqrt(FGm12)~s(DiaPres,by=Treatment)+s(FGm0, by=Treatment), data = hirs)
summary(mod1sqrts.4)
par(mfrow = c(2,2))
gam.check(mod1sqrts.4)
```

```{r}
anova(mod1sqrts.1, mod1sqrts.2,test='F')
```

As the anova is not significant the simpler model is better, so the model without the interaction.

For the following anova test the two additional models are tested for sake of argumentation but the results are no better than models mode1sqrts.1 and mod1sqrts.2.
```{r}
anova(mod1sqrts.3, mod1sqrts.4,test='F')
```

```{r}
mod2sqrts.1 <- gam(sqrt(FGm12)~te(FGm0,DiaPres) + Treatment, data = hirs)
summary(mod2sqrts.1)
```


```{r}
mod2sqrts.2 <- gam(sqrt(FGm12)~te(FGm0,SysPres) + Treatment , data = hirs)
summary(mod2sqrts.2)
```


```{r}
mod2sqrts.3 <- gam(sqrt(FGm12)~te(FGm0,weight) + Treatment, data = hirs)
summary(mod2sqrts.3)
```


```{r}
mod2sqrts.4 <- gam(sqrt(FGm12)~te(FGm0,height) + Treatment , data = hirs)
summary(mod2sqrts.4)
```

```{r}
AIC(mod1sqrts.1,mod2sqrts.1,mod2sqrts.2,mod2sqrts.3,mod2sqrts.4)
```
Out of all these new fitted models, the best one is the model with Treatment, and a tensor of FGm0 and height, as it is the one with the lowest AIC.
